# =============================================================================
# GED PERFORMANCE SEGMENTATION — K-MEDOIDS CLUSTERING
# =============================================================================
# Goal: Segment candidates using K-Medoids (handling mixed numeric/categorical).
# This script operates independently but maintains logic consistency with DIANA.
# =============================================================================

# import k prototypes packages
```{r}
required_pkgs <- c("tidyverse", "clustMixType", "scales", "knitr", "RColorBrewer")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
invisible(lapply(required_pkgs, library, character.only = TRUE))
```

# process preparation (follow DIANA's logic for consistency)
```{r}
set.seed(1234)

INPUT_FILE    <- "C://Users//amogh//Documents//GED_EDA_Project//data//processed//5_test_candidate_cleaned_final.csv"
VAR_THRESHOLD <- 0.15  # Clustering features below this variance are dropped
COR_THRESHOLD <- 0.80  # If two clustering features correlate above this, the lower-variance one is dropped to reduce redundancy
COMPLETENESS_THRESHOLD <- 0.70  # Candidates with fewer than this proportion of clustering fields present are removed before clustering to prevent missingness artifact clusters
```

# load data and build candidtate level dataset
```{r}
raw <- read_csv(INPUT_FILE, show_col_types = FALSE)
candidates <- raw %>%
  mutate(score_clean = if_else(SCORE_MISSING == 1L, NA_real_, SCORE)) %>%
  group_by(CANDIDATE_ID) %>%
  summarise(
    # --- Performance (used in clustering) ---
    avg_score          = mean(score_clean, na.rm = TRUE),
    pass_rate          = mean(if_else(!is.na(score_clean) & score_clean >= 145, 1L, 0L), na.rm = TRUE),
    college_ready_rate = mean(if_else(!is.na(score_clean) & score_clean >= 165, 1L, 0L), na.rm = TRUE),
    total_attempts     = n(),
    avg_retakes        = (n() / n_distinct(EXAM_SUBJECT)) - 1,
    credential_earned  = as.integer(any(!is.na(CREDENTIAL_DATE))),
    
    # Subject scores — NA when that subject was never attempted
    # Encoding: Math=1, Science=2, Reasoning=3, Social Studies=4
    score_math           = mean(score_clean[EXAM_SUBJECT == 1], na.rm = TRUE),
    score_science        = mean(score_clean[EXAM_SUBJECT == 2], na.rm = TRUE),
    score_reasoning      = mean(score_clean[EXAM_SUBJECT == 3], na.rm = TRUE),
    score_social_studies = mean(score_clean[EXAM_SUBJECT == 4], na.rm = TRUE),
    
    # --- Delivery (used in clustering) ---
    pct_online     = mean(as.integer(ON_VUE == 1), na.rm = TRUE),
    used_ged_ready = as.integer(any(GED_READY == 1, na.rm = TRUE)),
    has_prep_center = as.integer(first(PREP_CENTER) != "UNKNOWN"),
    
    # --- Preparation (not used in clustering) ---
    studied_for_ged          = first(STUDIED_FOR_GED),
    # Encoding: Work=1, Ed. Gain=2, Personal=3, Military=4, Special=5, Unknown=0
    testing_reason           = first(TESTING_REASON),
    # Encoding: Academic=1, Personal=2, Both=3, Neither=4,
    #           Homeschool=5, Foreign Diploma=6, Unknown=0
    school_incomplete_reason = first(SCHOOL_INCOMPLETE_REASON),
    # Encoding: Interested=1, Contacted=2, Enrolled=3,
    #           Credentialed=4, Dismissed=5, Unknown=0
    enrollment_status        = first(ENROLLMENT_STATUS),
    has_prep_center          = as.integer(first(PREP_CENTER) != "UNKNOWN"),
    study_adult_ed_class     = first(STUDY_HELPFUL_ADULT_EDUCATION_CLASS),
    study_books              = first(STUDY_HELPFUL_BOOKS_PRINTED_STUDY_MATERIAL),
    study_online_video       = first(STUDY_HELPFUL_ONLINE_COURSE_VIDEO_STUDY_MATERIALS),
    
    # --- Demographics (analysis only — NOT used in clustering) ---
    # Encoding: Male=1, Female=2, Nonbinary=3, Decline=4, Unknown=0
    gender           = first(GENDER),
    # Encoding: English=1, Spanish=2, Unknown=0
    language_code    = first(LANGUAGE_CODE),
    # Encoding: Under $5K=1 ... $75K+=8, Unknown=0
    last_year_income = first(LAST_YEAR_INCOME),
    # Encoding: PreK-5=1, 6-8=2, 9th=3, 10th=4, 11th=5, 12th=6, Don't Remember=7, Other=8, Never=9, Unknown=0
    highest_grade    = first(HIGHEST_GRADE_COMPLETED),
    # Encoding: Non-Hispanic=1, Hispanic=2, Decline=3, Unknown=0
    ethnicity        = first(ETHNICITY),
    race_white       = first(WHITE),
    race_black       = first(AFRICAN_AMERICAN),
    race_asian       = first(ASIAN),
    race_native      = first(INDIAN_OR_ALASKAN),
    race_pacific     = first(HAWAIIAN_OR_PACIFIC),
    race_declined    = first(RACE_DECLINE),
    race_none        = first(RACE_NONE),
    birth_year       = first(BIRTH_YEAR),
    
    # --- Geography (analysis only — NOT used in clustering) ---
    c_state = first(C_STATE),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(c(score_math, score_science, score_reasoning, score_social_studies),
           ~ if_else(is.nan(.x), NA_real_, .x)),
    
    # --- Derived demographic labels (analysis only) ---
    age = 2024 - birth_year,
    
    census_region = case_when(
      c_state %in% c("CT","ME","MA","NH","RI","VT","NJ","NY","PA") ~ "Northeast",
      c_state %in% c("IL","IN","MI","OH","WI","IA","KS","MN",
                     "MO","NE","ND","SD")                          ~ "Midwest",
      c_state %in% c("DE","FL","GA","MD","NC","SC","VA","WV","DC",
                     "AL","KY","MS","TN","AR","LA","OK","TX")      ~ "South",
      c_state %in% c("AZ","CO","ID","MT","NV","NM","UT","WY",
                     "AK","CA","HI","OR","WA")                     ~ "West",
      TRUE                                                          ~ "Other"
    ),
    
    race_category = case_when(
      race_declined == 1               ~ "Declined",
      race_none     == 1               ~ "Not Reported",
      race_black == 1 & ethnicity == 2 ~ "Black Hispanic",
      race_black == 1                  ~ "Black Non-Hispanic",
      ethnicity  == 2                  ~ "Hispanic",
      race_white == 1                  ~ "White Non-Hispanic",
      race_asian == 1                  ~ "Asian",
      race_native == 1                 ~ "Native American",
      race_pacific == 1                ~ "Pacific Islander",
      TRUE                             ~ "Multiracial/Other"
    ),
    
    gender_label = case_when(
      gender == 1 ~ "Male",      gender == 2 ~ "Female",
      gender == 3 ~ "Nonbinary", gender == 4 ~ "Decline",
      TRUE ~ "Unknown"
    ),
    language_label = case_when(
      language_code == 1 ~ "English", language_code == 2 ~ "Spanish",
      TRUE ~ "Unknown"
    ),
    income_label = case_when(
      last_year_income == 1 ~ "Under $5K",    last_year_income == 2 ~ "$5K-$9,999",
      last_year_income == 3 ~ "$10K-$19,999", last_year_income == 4 ~ "$20K-$29,999",
      last_year_income == 5 ~ "$30K-$39,999", last_year_income == 6 ~ "$40K-$49,999",
      last_year_income == 7 ~ "$50K-$74,999", last_year_income == 8 ~ "$75K+",
      TRUE ~ "Unknown"
    ),
    edlevel_label = case_when(
      highest_grade == 1 ~ "Pre-K to 5th",   highest_grade == 2 ~ "6th-8th",
      highest_grade == 3 ~ "9th Grade",      highest_grade == 4 ~ "10th Grade",
      highest_grade == 5 ~ "11th Grade",     highest_grade == 6 ~ "12th Grade",
      highest_grade == 7 ~ "Don't Remember", highest_grade == 8 ~ "Other",
      highest_grade == 9 ~ "Never Attended", TRUE ~ "Unknown"
    ),
    testing_reason_label = case_when(
      testing_reason == 1 ~ "Work",         testing_reason == 2 ~ "Education",
      testing_reason == 3 ~ "Personal",     testing_reason == 4 ~ "Military",
      testing_reason == 5 ~ "Special Req.", TRUE ~ "Unknown"
    ),
    school_incomplete_label = case_when(
      school_incomplete_reason == 1 ~ "Academic",
      school_incomplete_reason == 2 ~ "Personal",
      school_incomplete_reason == 3 ~ "Personal & Academic",
      school_incomplete_reason == 4 ~ "Neither",
      school_incomplete_reason == 5 ~ "Home Schooled",
      school_incomplete_reason == 6 ~ "Foreign Diploma",
      TRUE ~ "Unknown"
    ),
    enrollment_label = case_when(
      enrollment_status == 1 ~ "Interested", enrollment_status == 2 ~ "Contacted",
      enrollment_status == 3 ~ "Enrolled",   enrollment_status == 4 ~ "Credentialed",
      enrollment_status == 5 ~ "Dismissed",  TRUE ~ "Unknown"
    )
  )
```

# remove candidates with no demographic data
```{r}
candidates <- candidates %>%
  mutate(completeness = rowMeans(!is.na(select(., avg_score, pass_rate, pct_online, gender)))) %>%
  filter(completeness >= COMPLETENESS_THRESHOLD) %>%
  filter(!(gender == 0 & race_black == 0 & race_white == 0 & ethnicity == 0))
cat("Candidates after aggregation:", nrow(candidates), "\n")
```

# feature selection (follow DIANA's logic for consistency)
```{r}
# initial variables
candidate_vars <- c("avg_score", "pass_rate", "college_ready_rate", "avg_retakes", 
                   "pct_online", "credential_earned", "used_ged_ready", "has_prep_center",
                   "studied_for_ged", "study_books", "study_online_video")

# filtered by variance
numeric_data <- candidates %>% select(all_of(candidate_vars)) %>% select(where(is.numeric))
variances <- sapply(numeric_data, function(x) sd(x, na.rm=T) / abs(mean(x, na.rm=T)))
protected_var <- "avg_score"
keep_vars <- names(variances[variances >= VAR_THRESHOLD | names(variances) == protected_var])

# filter by correlation
cor_mat <- cor(candidates %>% select(all_of(keep_vars)), use = "pairwise.complete.obs")
high_cor_pairs <- which(abs(cor_mat) > COR_THRESHOLD & upper.tri(cor_mat), arr.ind = TRUE)

to_drop <- c()

if (nrow(high_cor_pairs) > 0) {
  for (i in 1:nrow(high_cor_pairs)) {
    var_a <- rownames(cor_mat)[high_cor_pairs[i, 1]]
    var_b <- colnames(cor_mat)[high_cor_pairs[i, 2]]
    
    # delete on variable if both are not already marked for deletion
    if (!(var_a %in% to_drop) & !(var_b %in% to_drop)) {
      # if one is avg_score then drop the other, otherwise drop the one with lower variance
      if (var_a == "avg_score") {
        to_drop <- c(to_drop, var_b)
      } else {
        to_drop <- c(to_drop, var_a)
      }
    }
  }
}
final_vars <- setdiff(keep_vars, to_drop)
drop_vars <- setdiff(candidate_vars, final_vars)

cat("Features deleted:", paste(drop_vars, collapse=", "), "\n")
cat("Remaining features:", paste(final_vars, collapse=", "), "\n")
```

# k-Medoids clustering
```{r}
cluster_df <- candidates %>%
  select(all_of(final_vars)) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, mean(.x, na.rm = TRUE))))
  
# standardize data
cluster_df_scaled <- cluster_df %>%
  mutate(across(where(is.numeric), ~ as.numeric(scale(.x)))) %>%
  
  # transfer binary data into factor for categorical variables
  mutate(across(any_of(c("credential_earned", "used_ged_ready", "has_prep_center", 
                         "studied_for_ged", "study_books", "study_online_video")), as.factor))

# Compute gower distance
library(cluster)

gower_dist <- daisy(cluster_df_scaled, metric = "gower")

# find the best k by Elbow Method
cat("\ncalculating elbow plot\n")

k_range <- 2:7
costs <- numeric(length(k_range))

for (i in seq_along(k_range)) {
  k <- k_range[i]
  temp_model <- pam(gower_dist, k = k, diss = TRUE)
  costs[i] <- temp_model$objective[1]
}

plot(k_range, costs, type = "b", pch = 19, col = "#2166AC",
     xlab = "Number of Clusters (K)",
     ylab = "Total Within-cluster Dissimilarity",
     main = "K-Medoids Elbow Plot")

```

# Silhoutte Analysis
```{r}
# --- Silhouette analysis: try k = 2 through 8 ---
cat("\nComputing silhouette widths for k = 2 through 8 (PAM)...\n")
cat("This may take a few minutes on large datasets.\n")

sil_results <- map_dfr(2:8, function(k) {
  pam_k  <- pam(gower_dist, k = k, diss = TRUE)
  tibble(k = k, avg_silhouette = round(mean(pam_k$silinfo$widths[, 3]), 4))
})

stat_opt <- sil_results$k[which.max(sil_results$avg_silhouette)]

sil_results %>%
  mutate(Note = if_else(k == stat_opt, "<-- statistical optimum", "")) %>%
  kable(format = "simple", col.names = c("k", "Avg Silhouette", "")) %>%
  print()

cat(sprintf("\nStatistical optimum: k = %d  |  Currently selected: k = %d\n",
            stat_opt, K))

# --- Silhouette plot ---
print(
  ggplot(sil_results, aes(x = k, y = avg_silhouette)) +
    geom_line(color = "#2166AC", linewidth = 1.3) +
    geom_point(size = 4.5, color = "#2166AC") +

    # Statistical optimum — grey diamond
    geom_point(data = filter(sil_results, k == stat_opt),
               size = 7, shape = 18, color = "grey30") +
    annotate("text",
             x     = stat_opt,
             y     = sil_results$avg_silhouette[sil_results$k == stat_opt],
             label = sprintf("Statistical optimum\n(k = %d, sil = %.3f)",
                             stat_opt,
                             sil_results$avg_silhouette[sil_results$k == stat_opt]),
             hjust = -0.1, vjust = 0.5, size = 3.2, color = "grey30") +

    # Selected K — red dot (only shown when different from stat_opt)
    { if (K != stat_opt) list(
      geom_point(data = filter(sil_results, k == K),
                 size = 7, shape = 21, fill = "#D6604D",
                 color = "white", stroke = 2),
      annotate("text",
               x     = K,
               y     = sil_results$avg_silhouette[sil_results$k == K],
               label = sprintf("Selected\n(k = %d, sil = %.3f)",
                               K,
                               sil_results$avg_silhouette[sil_results$k == K]),
               hjust = 1.1, vjust = -0.3, size = 3.2, color = "#D6604D"),
      annotate("rect",
               xmin = min(stat_opt, K) - 0.3, xmax = max(stat_opt, K) + 0.3,
               ymin = -Inf, ymax = Inf,
               fill = "#FEE8C8", alpha = 0.35)
    )
    } +

    scale_x_continuous(breaks = 2:8) +
    scale_y_continuous(labels = number_format(accuracy = 0.001)) +
    labs(
      title    = "Silhouette Analysis — How Many Clusters? (PAM)",
      subtitle = sprintf(
        "Grey diamond = statistical optimum (k = %d)%s\nHigher silhouette = better cluster separation",
        stat_opt,
        if (K != stat_opt) sprintf("  |  Red dot = selected k = %d", K) else ""
      ),
      x       = "Number of Clusters (k)",
      y       = "Average Silhouette Width",
      caption = "Tip: if k = 2 feels too coarse, look for the next local peak"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(color = "grey40", size = 9.5),
      plot.caption  = element_text(color = "grey50", size = 8.5, face = "italic")
    )
)

```
# Final Model
```{r}
K_FINAL <- 6
pam_k  <- pam(gower_dist, k = K_FINAL, diss = TRUE)

candidates <- candidates %>%
  mutate(cluster = pam_k$clustering)
```

# summary table for cluster profiles
```{r}
profile <- candidates %>%
  group_by(cluster) %>%
  summarise(
    n               = n(),
    avg_score       = round(mean(avg_score,          na.rm = TRUE), 1),
    pass_rate       = round(mean(pass_rate,          na.rm = TRUE), 3),
    college_ready   = round(mean(college_ready_rate, na.rm = TRUE), 3),
    credential_pct  = round(mean(credential_earned,  na.rm = TRUE), 3),
    avg_retakes     = round(mean(avg_retakes,        na.rm = TRUE), 2),
    pct_online      = round(mean(pct_online,         na.rm = TRUE), 3),
    pct_ged_ready   = round(mean(used_ged_ready,     na.rm = TRUE), 3),
    pct_prep_center = round(mean(has_prep_center,    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_score)) %>%
  mutate(perf_rank = row_number())
print(profile)
```

```{r}
# ── SECTION 7: NAME CLUSTERS ──────────────────────────────────────────────────
# Each cluster gets a short two-part name:
#   Part 1 — Performance tier (ranked by average score, 1 = highest)
#   Part 2 — The most distinctive preparation or delivery behavior for that
#             cluster, identified by the largest absolute z-score relative
#             to the sample mean across clusters.
#
# Tier assignments based on observed profiles (K = 6):
#   Rank 1 (avg 157.5) — Cluster 3: high pass rate, first-attempt, no support
#   Rank 2 (avg 156.3) — Cluster 4: high pass rate, retakers, 100% credential
#   Rank 3 (avg 154.4) — Cluster 2: retakers with GED Ready support
#   Rank 4 (avg 146.8) — Cluster 6: borderline passers, GED Ready users
#   Rank 5 (avg 145.8) — Cluster 5: borderline passers, prep center users
#   Rank 6 (avg 141.2) — Cluster 1: below threshold, no support, no credential

# Performance tier names — one per rank up to K
all_tier_names <- c(
  "High Achievers",          # rank 1: best scores
  "Strong Performers",       # rank 2
  "Solid Performers",        # rank 3
  "Developing Learners",     # rank 4
  "Below Average Learners",  # rank 5
  "Struggling Learners",     # rank 6
  "At-Risk Candidates",      # rank 7
  "Critical Needs"           # rank 8: lowest scores
)
tier_names <- all_tier_names[1:K_FINAL]

# Trait labels restricted to features that were actually in the clustering
# model. Study behaviors (studied_for_ged, adult ed, books, online video)
# are excluded — they are post-hoc observations and must not drive naming.
# Only pct_online, pct_ged_ready, pct_prep_center, and avg_retakes
# were clustering inputs alongside the performance outcome features.
trait_labels_pos <- c(
  pct_online      = "Online-Heavy",
  pct_ged_ready   = "GED Ready Users",
  pct_prep_center = "Prep Supported",
  avg_retakes     = "High Retakers"
)
trait_labels_neg <- c(
  pct_online      = "In-Person Testers",
  pct_ged_ready   = "No GED Ready",
  pct_prep_center = "No Prep Center",
  avg_retakes     = "First-Attempt Passers"
)

# Sample-wide averages as baseline for z-score comparison
# Only includes the four clustering features eligible for naming
benchmarks <- list(
  pct_online      = mean(candidates$pct_online,      na.rm = TRUE),
  pct_ged_ready   = mean(candidates$used_ged_ready,  na.rm = TRUE),
  pct_prep_center = mean(candidates$has_prep_center, na.rm = TRUE),
  avg_retakes     = mean(candidates$avg_retakes,     na.rm = TRUE)
)

cluster_names <- profile %>%
  mutate(
    tier  = tier_names[perf_rank],
    trait = map_chr(seq_len(n()), function(i) {
      row <- profile[i, ]
      z_scores <- map_dbl(names(trait_labels_pos), function(feat) {
        val  <- as.numeric(row[[feat]])
        avg  <- benchmarks[[feat]]
        sdev <- sd(profile[[feat]], na.rm = TRUE)
        if (is.na(sdev) | sdev == 0) return(0)
        (val - avg) / sdev
      })
      names(z_scores) <- names(trait_labels_pos)

      # Pick the feature with the largest absolute z-score
      best_feat <- names(which.max(abs(z_scores)))

      # Choose the positive or negative label based on direction of z-score
      if (z_scores[best_feat] >= 0) {
        trait_labels_pos[best_feat]
      } else {
        trait_labels_neg[best_feat]
      }
    }),
    cluster_name = paste0(tier, " — ", trait)
  ) %>%
  group_by(cluster_name) %>%
  mutate(cluster_name = if_else(
    n() > 1,
    paste0(cluster_name, " (", row_number(), ")"),
    cluster_name
  )) %>%
  ungroup() %>%
  select(cluster, cluster_name)

candidates <- left_join(candidates, cluster_names, by = "cluster")

# Print cluster name assignments for review
cat("\n--- Cluster Name Assignments ---\n")
candidates %>%
  count(cluster, cluster_name) %>%
  arrange(cluster) %>%
  kable(format = "simple", col.names = c("Cluster", "Name", "N")) %>%
  print()

# ── SECTION 8: CLUSTER SUMMARY ────────────────────────────────────────────────
# Tables 1–2 cover the features used in clustering (performance outcomes).
# Tables 3–4 are post-hoc profiles — these features did not influence the
# clusters, so any patterns found here are genuine discoveries:
#   Table 3 — Preparation behaviors (study methods, GED Ready, prep center)
#   Table 4 — Demographic composition (race, gender, language, geography)

cat("\n=== CLUSTER SUMMARY ===\n")

pct  <- function(x) percent(mean(x, na.rm = TRUE), 0.1)
avg  <- function(x) round(mean(x, na.rm = TRUE), 1)
avg2 <- function(x) round(mean(x, na.rm = TRUE), 2)

base_profile <- candidates %>%
  group_by(Cluster = cluster, Name = cluster_name) %>%
  summarise(N = n(), .groups = "drop") %>%
  arrange(Cluster)

# --- Table 1: Performance ---
cat("\n--- Performance (clustering features) ---\n")
candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    `Avg Score`         = avg(avg_score),
    `Pass Rate`         = pct(pass_rate),
    `College Ready`     = pct(college_ready_rate),
    `Credential Earned` = pct(credential_earned),
    `Total Attempts`    = avg2(total_attempts),
    `Avg Retakes`       = avg2(avg_retakes),
    .groups = "drop"
  ) %>%
  left_join(base_profile %>% select(Cluster, Name, N), by = "Cluster") %>%
  select(Cluster, Name, N, everything()) %>%
  arrange(Cluster) %>%
  kable(format = "simple") %>%
  print()

# --- Table 2: Subject Scores ---
cat("\n--- Subject Scores (clustering features | passing threshold = 145) ---\n")
candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    `Math`           = avg(score_math),
    `Science`        = avg(score_science),
    `Reasoning`      = avg(score_reasoning),
    `Social Studies` = avg(score_social_studies),
    .groups = "drop"
  ) %>%
  left_join(base_profile %>% select(Cluster, Name), by = "Cluster") %>%
  select(Cluster, Name, everything()) %>%
  arrange(Cluster) %>%
  kable(format = "simple") %>%
  print()

# --- Table 3: Preparation Behaviors ---
cat("\n--- Preparation Behaviors (post-hoc — not used in clustering) ---\n")
candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    `Studied for GED`     = pct(studied_for_ged),
    `Used GED Ready`      = pct(used_ged_ready),
    `Has Prep Center`     = pct(has_prep_center),
    `Online Testing`      = pct(pct_online),
    `Study: Adult Ed`     = pct(study_adult_ed_class),
    `Study: Books`        = pct(study_books),
    `Study: Online Video` = pct(study_online_video),
    .groups = "drop"
  ) %>%
  left_join(base_profile %>% select(Cluster, Name), by = "Cluster") %>%
  select(Cluster, Name, everything()) %>%
  arrange(Cluster) %>%
  kable(format = "simple") %>%
  print()

# --- Table 4: Demographic Profile (post-hoc — NOT used in clustering) ---
cat("\n--- Demographic Profile (post-hoc observation — not used in clustering) ---\n")
cat("    These patterns emerged from the data; they were not inputs to the model.\n\n")
candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    `Avg Age`          = avg(age),
    `Male`             = pct(gender == 1),
    `Female`           = pct(gender == 2),
    `Black`            = pct(race_black == 1),
    `White`            = pct(race_white == 1),
    `Hispanic`         = pct(ethnicity == 2),
    `Asian`            = pct(race_asian == 1),
    `Spanish Lang`     = pct(language_code == 2),
    `Top Region`       = names(sort(table(census_region), decreasing = TRUE))[1],
    `Top State`        = names(sort(table(c_state),       decreasing = TRUE))[1],
    `Income Unknown`   = pct(last_year_income == 0),
    `Ed: 9th or Below` = pct(highest_grade <= 3 & highest_grade > 0),
    .groups = "drop"
  ) %>%
  left_join(base_profile %>% select(Cluster, Name), by = "Cluster") %>%
  select(Cluster, Name, everything()) %>%
  arrange(Cluster) %>%
  kable(format = "simple") %>%
  print()

cat(sprintf(
  "\nPAM complete | Avg Silhouette = %.4f | k = %d | Clustering features = %d\n",
  mean(pam_k$silinfo$widths[, 3]), K_FINAL,
  length(retained_numeric) + length(retained_nominal)
))
cat("Demographics are post-hoc only and were not used in clustering.\n")
```

