# =============================================================================
# GED PERFORMANCE SEGMENTATION — K-PROTOTYPES CLUSTERING
# =============================================================================
# Goal: Segment candidates using K-Prototypes (handling mixed numeric/categorical).
# This script operates independently but maintains logic consistency with DIANA.
# =============================================================================

# import k prototypes packages
```{r}
required_pkgs <- c("tidyverse", "clustMixType", "scales", "knitr", "RColorBrewer")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
invisible(lapply(required_pkgs, library, character.only = TRUE))
```

# process preparation (follow DIANA's logic for consistency)
```{r}
set.seed(1234)

INPUT_FILE    <- "/Users/chentzuyu/Desktop/study_abroad/course/spring/6411_Exploratory_Data_Analytics/6411_group/5_test_candidate_cleaned_final.csv"
VAR_THRESHOLD <- 0.15  # Clustering features below this variance are dropped
COR_THRESHOLD <- 0.80  # If two clustering features correlate above this, the lower-variance one is dropped to reduce redundancy
COMPLETENESS_THRESHOLD <- 0.70  # Candidates with fewer than this proportion of clustering fields present are removed before clustering to prevent missingness artifact clusters
```

# load data and build candidtate level dataset
```{r}
raw <- read_csv(INPUT_FILE, show_col_types = FALSE)
candidates <- raw %>%
  mutate(score_clean = if_else(SCORE_MISSING == 1L, NA_real_, SCORE)) %>%
  group_by(CANDIDATE_ID) %>%
  summarise(
    # --- Performance (used in clustering) ---
    avg_score          = mean(score_clean, na.rm = TRUE),
    pass_rate          = mean(if_else(!is.na(score_clean) & score_clean >= 145, 1L, 0L), na.rm = TRUE),
    college_ready_rate = mean(if_else(!is.na(score_clean) & score_clean >= 165, 1L, 0L), na.rm = TRUE),
    total_attempts     = n(),
    avg_retakes        = (n() / n_distinct(EXAM_SUBJECT)) - 1,
    credential_earned  = as.integer(any(!is.na(CREDENTIAL_DATE))),
    
    # Subject scores — NA when that subject was never attempted
    # Encoding: Math=1, Science=2, Reasoning=3, Social Studies=4
    score_math           = mean(score_clean[EXAM_SUBJECT == 1], na.rm = TRUE),
    score_science        = mean(score_clean[EXAM_SUBJECT == 2], na.rm = TRUE),
    score_reasoning      = mean(score_clean[EXAM_SUBJECT == 3], na.rm = TRUE),
    score_social_studies = mean(score_clean[EXAM_SUBJECT == 4], na.rm = TRUE),
    
    # --- Delivery (used in clustering) ---
    pct_online     = mean(as.integer(ON_VUE == 1), na.rm = TRUE),
    used_ged_ready = as.integer(any(GED_READY == 1, na.rm = TRUE)),
    has_prep_center = as.integer(first(PREP_CENTER) != "UNKNOWN"),
    
    # --- Preparation (not used in clustering) ---
    studied_for_ged          = first(STUDIED_FOR_GED),
    # Encoding: Work=1, Ed. Gain=2, Personal=3, Military=4, Special=5, Unknown=0
    testing_reason           = first(TESTING_REASON),
    # Encoding: Academic=1, Personal=2, Both=3, Neither=4,
    #           Homeschool=5, Foreign Diploma=6, Unknown=0
    school_incomplete_reason = first(SCHOOL_INCOMPLETE_REASON),
    # Encoding: Interested=1, Contacted=2, Enrolled=3,
    #           Credentialed=4, Dismissed=5, Unknown=0
    enrollment_status        = first(ENROLLMENT_STATUS),
    has_prep_center          = as.integer(first(PREP_CENTER) != "UNKNOWN"),
    study_adult_ed_class     = first(STUDY_HELPFUL_ADULT_EDUCATION_CLASS),
    study_books              = first(STUDY_HELPFUL_BOOKS_PRINTED_STUDY_MATERIAL),
    study_online_video       = first(STUDY_HELPFUL_ONLINE_COURSE_VIDEO_STUDY_MATERIALS),
    
    # --- Demographics (analysis only — NOT used in clustering) ---
    # Encoding: Male=1, Female=2, Nonbinary=3, Decline=4, Unknown=0
    gender           = first(GENDER),
    # Encoding: English=1, Spanish=2, Unknown=0
    language_code    = first(LANGUAGE_CODE),
    # Encoding: Under $5K=1 ... $75K+=8, Unknown=0
    last_year_income = first(LAST_YEAR_INCOME),
    # Encoding: PreK-5=1, 6-8=2, 9th=3, 10th=4, 11th=5, 12th=6, Don't Remember=7, Other=8, Never=9, Unknown=0
    highest_grade    = first(HIGHEST_GRADE_COMPLETED),
    # Encoding: Non-Hispanic=1, Hispanic=2, Decline=3, Unknown=0
    ethnicity        = first(ETHNICITY),
    race_white       = first(WHITE),
    race_black       = first(AFRICAN_AMERICAN),
    race_asian       = first(ASIAN),
    race_native      = first(INDIAN_OR_ALASKAN),
    race_pacific     = first(HAWAIIAN_OR_PACIFIC),
    race_declined    = first(RACE_DECLINE),
    race_none        = first(RACE_NONE),
    birth_year       = first(BIRTH_YEAR),
    
    # --- Geography (analysis only — NOT used in clustering) ---
    c_state = first(C_STATE),
    
    .groups = "drop"
  ) %>%
  mutate(
    across(c(score_math, score_science, score_reasoning, score_social_studies),
           ~ if_else(is.nan(.x), NA_real_, .x)),
    
    # --- Derived demographic labels (analysis only) ---
    age = 2024 - birth_year,
    
    census_region = case_when(
      c_state %in% c("CT","ME","MA","NH","RI","VT","NJ","NY","PA") ~ "Northeast",
      c_state %in% c("IL","IN","MI","OH","WI","IA","KS","MN",
                     "MO","NE","ND","SD")                          ~ "Midwest",
      c_state %in% c("DE","FL","GA","MD","NC","SC","VA","WV","DC",
                     "AL","KY","MS","TN","AR","LA","OK","TX")      ~ "South",
      c_state %in% c("AZ","CO","ID","MT","NV","NM","UT","WY",
                     "AK","CA","HI","OR","WA")                     ~ "West",
      TRUE                                                          ~ "Other"
    ),
    
    race_category = case_when(
      race_declined == 1               ~ "Declined",
      race_none     == 1               ~ "Not Reported",
      race_black == 1 & ethnicity == 2 ~ "Black Hispanic",
      race_black == 1                  ~ "Black Non-Hispanic",
      ethnicity  == 2                  ~ "Hispanic",
      race_white == 1                  ~ "White Non-Hispanic",
      race_asian == 1                  ~ "Asian",
      race_native == 1                 ~ "Native American",
      race_pacific == 1                ~ "Pacific Islander",
      TRUE                             ~ "Multiracial/Other"
    ),
    
    gender_label = case_when(
      gender == 1 ~ "Male",      gender == 2 ~ "Female",
      gender == 3 ~ "Nonbinary", gender == 4 ~ "Decline",
      TRUE ~ "Unknown"
    ),
    language_label = case_when(
      language_code == 1 ~ "English", language_code == 2 ~ "Spanish",
      TRUE ~ "Unknown"
    ),
    income_label = case_when(
      last_year_income == 1 ~ "Under $5K",    last_year_income == 2 ~ "$5K-$9,999",
      last_year_income == 3 ~ "$10K-$19,999", last_year_income == 4 ~ "$20K-$29,999",
      last_year_income == 5 ~ "$30K-$39,999", last_year_income == 6 ~ "$40K-$49,999",
      last_year_income == 7 ~ "$50K-$74,999", last_year_income == 8 ~ "$75K+",
      TRUE ~ "Unknown"
    ),
    edlevel_label = case_when(
      highest_grade == 1 ~ "Pre-K to 5th",   highest_grade == 2 ~ "6th-8th",
      highest_grade == 3 ~ "9th Grade",      highest_grade == 4 ~ "10th Grade",
      highest_grade == 5 ~ "11th Grade",     highest_grade == 6 ~ "12th Grade",
      highest_grade == 7 ~ "Don't Remember", highest_grade == 8 ~ "Other",
      highest_grade == 9 ~ "Never Attended", TRUE ~ "Unknown"
    ),
    testing_reason_label = case_when(
      testing_reason == 1 ~ "Work",         testing_reason == 2 ~ "Education",
      testing_reason == 3 ~ "Personal",     testing_reason == 4 ~ "Military",
      testing_reason == 5 ~ "Special Req.", TRUE ~ "Unknown"
    ),
    school_incomplete_label = case_when(
      school_incomplete_reason == 1 ~ "Academic",
      school_incomplete_reason == 2 ~ "Personal",
      school_incomplete_reason == 3 ~ "Personal & Academic",
      school_incomplete_reason == 4 ~ "Neither",
      school_incomplete_reason == 5 ~ "Home Schooled",
      school_incomplete_reason == 6 ~ "Foreign Diploma",
      TRUE ~ "Unknown"
    ),
    enrollment_label = case_when(
      enrollment_status == 1 ~ "Interested", enrollment_status == 2 ~ "Contacted",
      enrollment_status == 3 ~ "Enrolled",   enrollment_status == 4 ~ "Credentialed",
      enrollment_status == 5 ~ "Dismissed",  TRUE ~ "Unknown"
    )
  )
```

# remove candidates with no demographic data
```{r}
candidates <- candidates %>%
  mutate(completeness = rowMeans(!is.na(select(., avg_score, pass_rate, pct_online, gender)))) %>%
  filter(completeness >= COMPLETENESS_THRESHOLD) %>%
  filter(!(gender == 0 & race_black == 0 & race_white == 0 & ethnicity == 0))
cat("Candidates after aggregation:", nrow(candidates), "\n")
```

# feature selection (follow DIANA's logic for consistency)
```{r}
# initial variables
candidate_vars <- c("avg_score", "pass_rate", "college_ready_rate", "avg_retakes", 
                   "pct_online", "credential_earned", "used_ged_ready", "has_prep_center",
                   "studied_for_ged", "study_books", "study_online_video")

# filtered by variance
numeric_data <- candidates %>% select(all_of(candidate_vars)) %>% select(where(is.numeric))
variances <- sapply(numeric_data, function(x) sd(x, na.rm=T) / abs(mean(x, na.rm=T)))
protected_var <- "avg_score"
keep_vars <- names(variances[variances >= VAR_THRESHOLD | names(variances) == protected_var])

# filter by correlation
cor_mat <- cor(candidates %>% select(all_of(keep_vars)), use = "pairwise.complete.obs")
high_cor_pairs <- which(abs(cor_mat) > COR_THRESHOLD & upper.tri(cor_mat), arr.ind = TRUE)

to_drop <- c()

if (nrow(high_cor_pairs) > 0) {
  for (i in 1:nrow(high_cor_pairs)) {
    var_a <- rownames(cor_mat)[high_cor_pairs[i, 1]]
    var_b <- colnames(cor_mat)[high_cor_pairs[i, 2]]
    
    # delete on variable if both are not already marked for deletion
    if (!(var_a %in% to_drop) & !(var_b %in% to_drop)) {
      # if one is avg_score then drop the other, otherwise drop the one with lower variance
      if (var_a == "avg_score") {
        to_drop <- c(to_drop, var_b)
      } else {
        to_drop <- c(to_drop, var_a)
      }
    }
  }
}
final_vars <- setdiff(keep_vars, to_drop)
drop_vars <- setdiff(candidate_vars, final_vars)

cat("Features deleted:", paste(drop_vars, collapse=", "), "\n")
cat("Remaining features:", paste(final_vars, collapse=", "), "\n")
```

# k-prototypes clustering
# k_prototypes metrics
```{r}
cluster_df <- candidates %>%
  select(all_of(clustering_vars)) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, mean(.x, na.rm = TRUE))))
  
# standardize data
cluster_df_scaled <- cluster_df %>%
  mutate(across(where(is.numeric), ~ as.numeric(scale(.x)))) %>%
  
  # transfer binary data into factor for categorical variables
  mutate(across(any_of(c("credential_earned", "used_ged_ready", "has_prep_center", 
                         "studied_for_ged", "study_books", "study_online_video")), as.factor))

# find the best k by Elbow Method
cat("\ncalculating elbow plot\n")

k_range <- 2:7
costs <- numeric(length(k_range))

for(i in seq_along(k_range)) {
  k <- k_range[i]
  temp_model <- kproto(cluster_df_scaled, k = k, nstart = 15, verbose = FALSE)
  costs[i] <- temp_model$tot.withinss
}

plot(k_range, costs, type = "b", pch = 19, col = "#2166AC",
     xlab = "Number of Clusters (K)", 
     ylab = "Total Within-cluster Cost",
     main = "K-Prototypes Elbow Plot")
```
```{r}
# modeling
K_FINAL <- 5
model <- kproto(cluster_df_scaled, k = K_FINAL, nstart = 5)

# clustering result
candidates$cluster <- as.factor(model$cluster)

```

```{r}
# summary table for cluster profiles
summary_table <- candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    N = n(),
    Score = round(mean(avg_score, na.rm = TRUE), 1),
    Pass = percent(mean(pass_rate), 0.1),
    Online = percent(mean(pct_online), 0.1),
    Female = percent(mean(gender == 2)),
    .groups = "drop"
  ) %>% arrange(desc(Score))

print(kable(summary_table, caption = "K-Prototypes Final Cluster Profiles"))
```

# Demorgraphic profiling
```{r}
# Demographic Profiling

demographic_profile <- candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    N = n(),
    Avg_Score  = round(mean(avg_score, na.rm = TRUE), 1),
    
    Avg_Age    = round(mean(age, na.rm = TRUE), 1),
    Pct_Female = percent(mean(gender == 2, na.rm = TRUE), 0.1),
    
    Pct_White    = percent(mean(race_white == 1, na.rm = TRUE), 0.1),
    Pct_Black    = percent(mean(race_black == 1, na.rm = TRUE), 0.1),
    Pct_Asian    = percent(mean(race_asian == 1, na.rm = TRUE), 0.1),
    Pct_Native   = percent(mean(race_native == 1, na.rm = TRUE), 0.1),
    Pct_Pacific  = percent(mean(race_native == 1, na.rm = TRUE), 0.1),
    Pct_Hispanic = percent(mean(race_pacific == 2, na.rm = TRUE), 0.1),
    
    Pct_High_Income = percent(mean(last_year_income >= 7, na.rm = TRUE), 0.1), # above $50K
    Pct_Low_Income  = percent(mean(last_year_income <= 2, na.rm = TRUE), 0.1), # below $10K 
    
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Score))

print(kable(demographic_profile, caption = "Demographic Analysis by Performance Cluster"))

```

```{r}
# Clusters by Geography

full_profile <- candidates %>%
  group_by(Cluster = cluster) %>%
  summarise(
    N = n(),
    Avg_Score = round(mean(avg_score, na.rm = TRUE), 1),
    
    Pct_South     = percent(mean(census_region == "South", na.rm = TRUE), 0.1),
    Pct_Midwest   = percent(mean(census_region == "Midwest", na.rm = TRUE), 0.1),
    Pct_Northeast = percent(mean(census_region == "Northeast", na.rm = TRUE), 0.1),
    Pct_West      = percent(mean(census_region == "West", na.rm = TRUE), 0.1),
    
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Score))

print(kable(full_profile, caption = "Regional and Demographic Analysis by Cluster"))

```



